Miniproject - Algebraic Geometry

### 2021-12-14 Algebraic geometry/submission.tex
Fixed alg geo, hopefully finally

    Here $\mathrm{lt}(I)$ is the leading term ideal of $I$, i.e. the ideal generated by the leading terms of all $f \in I$, w.r.t. $\preceq$.
\end{proposition}
\begin{proof}
    See \cite{kreuzer}.
\end{proof}
To apply this, first of all we have to collect information about the leading term ideal of $\I(\Gr(2, n))$.
This is done in the following lemma.
\begin{lemma}
    \label{prop:leading_term_ideal_grassmanian}
    Define the graded reverse monomial ordering $\preceq$ on $R := k[x_{i, j} \ | \ i < j]$ where the variables $x_{i, j}$ are ordered co-lexicographically w.r.t. $(i, j)$, i.e.
    \begin{equation*}
        x_{i, j} \leq x_{u, v} \ :\Leftrightarrow \ (i, j) \leq_{\text{colex}} (u, v)
    \end{equation*}
    Moreover, let
    \begin{equation*}
        I := \langle x_{i, j}x_{u, v} + x_{i, v}x_{j, u} - x_{i, u}x_{j, v} \ | \ i < j < u < v \rangle \leq R
    \end{equation*}
    be the ideal defining $\phi(\Gr(d, V))$ that was considered above. Then
    \begin{equation*}
        \mathrm{lt}(I) = J := \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle \leq R
    \end{equation*}
\end{lemma}
\begin{proof}
    Note that for $i < j < u < v$ have $x_{u, v}, x_{j, v} \succ x_{i, v} \succ x_{i, j}, x_{i, u}, x_{j, u}$.
    Thus the leading term of $x_{i, j} x_{u, v} + x_{i, v} x_{j, u} - x_{i, u} x_{j, v}$ is
    \begin{equation*}
        \mathrm{lt}(x_{i, j} x_{u, v} + x_{i, v} x_{j, u} - x_{i, u} x_{j, v}) = x_{i, v} x_{j, u}
    \end{equation*}
    It follows that $J \subseteq \mathrm{lt}(I)$.

    For the other direction, we use a quite lengthy degree argument.
    Sadly, the only argument I came up with is extremely technical.
    We try to present it as clearly as possible, at the cost of only sketching some parts.
    In the end, I checked it using Computer Algebra, and everything fits together.

    Consider an arbitrary element $F$ of $I$, given by homogeneous polynomials $f_{i, j, u, v} \in R$ via
    \begin{equation*}
        F = \sum_{i < j < u < v} f_{i, j, u, v} \ (x_{i, v} x_{j, u} - x_{i, u} x_{j, v} + x_{i, j} x_{u, v})
    \end{equation*}
    We want to show that $\mathrm{lt}(F) \in J$.

    Let
    \begin{equation*}
        f_{i, j, u, v} = \sum_{\alpha \in \N^N} c^{(i, j, u, v)}_\alpha x^\alpha
    \end{equation*}
    Then
    \begin{equation*}
        F = \sum_{\alpha \in \N^N} \sum_{i < j < u < v} c^{(i, j, u, v)}_\alpha x^\alpha \ (x_{i, v} x_{j, u} - x_{i, u} x_{j, v} + x_{i, j} x_{u, v})
    \end{equation*}
    and so there exists $\alpha \in \N^N$ and $\epsilon \in k^*$ with
    \begin{equation*}
        \mathrm{lt}(F) = \epsilon \cdot \mathrm{lt}\Bigl( \sum_{i < j < u < v} c_\alpha^{(i, j, u, v)} x^\alpha \ (x_{i, v} x_{j, u} - x_{i, u} x_{j, v} + x_{i, j} x_{u, v}) \Bigr)
    \end{equation*}
    Hence, we may assume wlog that all the $f_{i, j, u, v}$ are scaled monomials.

    Now, observe that all monomials in $F$ are of the form
    \begin{align*}
        \epsilon \ x_{i, v} \ x_{j, u} \ f_{i, j, u, v} \quad \text{or} \quad 
        \epsilon \ x_{i, u} \ x_{j, v} \ f_{i, j, u, v} \quad \text{or} \quad 
        \epsilon \ x_{i, j} \ x_{u, v} \ f_{i, j, u, v} 
    \end{align*}
    where $\epsilon \in k^*$ and $i < j < u < v$.
    In particular, this is true for the leading term $\mathrm{lt}(F)$.
    In the first of those case, clearly $\mathrm{lt}(F) \in J$.

    So consider now the second case, i.e. $\mathrm{lt}(F) = \epsilon \ x_{i, u} \ x_{j, v} \ f_{i, j, u, v}$.
    Since $x_{i, v} \ x_{j, u} \ f_{i, j, u, v} \succ x_{i, u} \ x_{j, v} \ f_{i, j, u, v}$, we see that the term $x_{i, v} \ x_{j, u} \ f_{i, j, u, v}$ cannot occur in $F$, i.e. must ``cancel out''.
    Hence the monomial $\epsilon \ x_{i, v} \ x_{j, u} \ f_{i, j, u, v}$ has a nonzero coefficient in
    \begin{equation*}
        F - f_{i, j, u, v} \ x_{i, v} \ x_{j, u} = \sum_{\substack{a < b < c < d\\(a, b, c, d) \neq (i, j, u, v)}} f_{a, b, c, d} \ (x_{a, d} x_{b, c} - x_{a, c} x_{b, d} + x_{a, b} x_{c, d})
    \end{equation*}
    and so for $(a, b, c, d) \neq (i, j, u, v), \epsilon' \in k^*$ have that
    \begin{align*}
        f_{i, j, u, v} \ x_{i, v} \ x_{j, u} &= \epsilon' f_{a, b, c, d} \ x_{a, d} \ x_{b, c} \quad \text{or}\\
        f_{i, j, u, v} \ x_{i, v} \ x_{j, u} &= \epsilon' f_{a, b, c, d} \ x_{a, c} \ x_{b, d} \quad \text{or}\\
        f_{i, j, u, v} \ x_{i, v} \ x_{j, u} &= \epsilon' f_{a, b, c, d} \ x_{a, b} \ x_{c, d}
    \end{align*}
    However, the second and third case imply that $f_{a, b, c, d} \ x_{a, d} \ x_{b, c} \succ f_{i, j, u, v} \ x_{i, v} \ x_{j, u}$.
    Hence, the ``new'' monomial $f_{a, b, c, d} \ x_{a, d} \ x_{b, c}$ also has to ``cancel out'' in the sum representation of $F$, as it is greater than $\mathrm{lt}(F) = \epsilon x_{i, u} \ x_{j, v} \ f_{i, j, u, v}$ in the order $\preceq$.
    So applying the whole argument inductively (induction on the number of monomials $\succ f_{i, j, u, v} \ x_{i, v} \ x_{j, u}$ that occur in any of the polynomials we work with), we end up in the first case
    \footnote{There is a small argument missing here, namely that we apply the next argument on each step of the induction, to show the claim for $f_{i, j, u, v} \ x_{i, v} \ x_{j, u}$. However, it should be easy to see that this is possible.}.

    If $(a, d) = (i, v)$ then $(b, c) \neq (j, u)$ and $x_{b, c} \divides f_{i, j, u, v}$.
    Thus
    \begin{equation*}
        x_{b, c} \ x_{i, u} \ x_{j, v} \divides \mathrm{lt}(F) = \epsilon \ x_{i, u} \ x_{j, v} \ f_{i, j, u, v}
    \end{equation*}
    with $i < j < u < v$ and $i < b < c < v$. 
    No matter how $j, u, b, c$ are ordered relatively to each other, we see that in each possible case $x_{b, c} \ x_{i, u} \ x_{j, v} \in J$:
    \begin{align*}
        j < u < b < c \ &\Rightarrow \ x_{j, v} x_{b, c} \in J \\
        j < b < u < c \ &\Rightarrow \ x_{j, v} x_{b, c} \in J \\
        j < b < c < u \ &\Rightarrow \ x_{i, u} x_{b, c} \in J \\
        b < j < u < c \ &\Rightarrow \ f_{i, b, c, v} \ x_{i, c} \ x_{b, v} \succ f_{i, j, u, v} \ x_{i, u} \ x_{j, v}, \ \contradiction \\
        b < j < c < u \ &\Rightarrow \ x_{i, u} x_{b, c} \in J \\
        b < c < j < u \ &\Rightarrow \ x_{i, u} x_{b, c} \in J \\
    \end{align*}
    Hence $\mathrm{lt}(F) \in J$.

    It is now left to consider the case $(b, c) = (j, u)$ and the case $(a, d) \neq (i, v), (b, c) \neq (j, u)$.
    The former can be dealt with in exactly the same way, by noting that $x_{a, d} \divides f_{i, j, u, v}$.
    In the latter case, we even find $x_{a, d} \ x_{b, c} \divides f_{i, j, u, v}$ and a very similar argument works.

    Finally, one must also consider the third ``big'' case, namely that
    \begin{equation*}
        \mathrm{lt}(F) = \epsilon \ x_{i, j} \ x_{u, v} \ f_{i, j, u, v} 
    \end{equation*}
    Again, you can do this similarly as before, but now two monomials ``cancel out''.
    We will not present this here as well.
\end{proof}

\begin{definition}
    Consider the graph $G_n = (V_n, E_n)$ where $V_n = \{ (i, j) \ | \ 0 \leq i < j \leq n \}$ and
    \begin{equation*}
        E_n = \{ \{ (i, v), (j, u) \} \ | \ 0 \leq i < j < u < v \leq n \}
    \end{equation*}
    Let further
    \begin{equation*}
        s_n(d) := |\{ I \subseteq V_n \ | \ I \ \text{independent set of size $d$}\}|
    \end{equation*}
\end{definition}

\begin{proposition}
    Let $d$ be the largest integer such that $s_n(d) \neq 0$, i.e. the size of the largest independent set in $G_n$.
    Then
    \begin{equation*}
        \dim(\Gr(2, n)) = d - 1
    \end{equation*}
    and
    \begin{equation*}
        \deg(\Gr(2, n)) = s_n(d)
    \end{equation*}
\end{proposition}
\begin{proof}
    The idea is to apply Macaulay's basis theorem to get information about the Hilbert function of $\Gr(2, n)$.
    To this end, observe that all the monomials generating $\mathrm{lt}(\I(\Gr(2, n)))$ are square-free.
    Hence, we can partition all monomials not in $\mathrm{lt}(\I(\Gr(2, n)))$ into sets depending on which variables occur in them.
    Doing this shows that leading coefficient of the Hilbert polynomial depends only on the groups with the maximal count of monomials.
    These monomials then form a maximal independent set in $G_n$, which allows us to relate degree and dimension of $\Gr(2, n)$ to $s_n(d)$.

@@

    On the other hand, it is easy to observe that the 0-th layer
    \begin{equation*}
        I = \{ v \in V \ | \ l(v) = 0 \}
    \end{equation*}
    is indeed an independent set, as for all elements $(i, j) \in I$ have either $i = 1$ or $j = n$.
\end{proof}
\begin{corollary}
    Have $\dim(\Gr(2, n)) = 2n - 4$.
\end{corollary}
\begin{proof}
    Follows directly from the previous two statements.
\end{proof}
I also think that it is not too difficult to show that
\begin{equation*}
    s_n(2n - 3) = \frac {(2(n - 2))!} {(n - 2)! (n - 1)!}
\end{equation*}
which then shows the degree formula for the case $d = 2$.
However, this is not a lecture on graph theory or combinatorics, and I already spent a lot of hours on this problem (and there are other mini projects as well :)).
So I will not do the proof here, but I think it is already very interesting to see those graphs and the connection of $\deg(\Gr(2, n))$ to combinatorial problems.
Finally, in the case of $n = 6$ we can simply count the independent sets by hand, and find the following result.
\begin{corollary}[Question (e)]
    Have $\dim(\Gr(2, 6)) = 8$ and $\deg(\Gr(2, 6)) = 14$.
\end{corollary}
\begin{proof}
    The dimension follows directly from the previous general case.
    For the degree, just count the maximal independent sets in $G_6$, as displayed in Example~\ref{ex:Gn}.
    This is slightly tricky, but it is not too hard to see that there are exactly 14 of them.
\end{proof}
Of course, the above statement on $\Gr(2, 6)$ can also easily be found using Computer Algebra.
For example, the following Sage script shows that
\begin{equation*}
    p_{\Gr(2, 6)} = \frac 1 {2880} t^8 + \frac 1 {120} t^7 + \frac {41} {480} t^6 + \frac {39} {80} t^5 + \frac {541} {320} t^4 + \frac {291} {80} t^3 + \frac {3401} {720} t^2 + \frac {101} {30} t + 1
\end{equation*}
from which we easily deduce that
\begin{equation*}
    \dim(\Gr(2, 6)) = 8 \quad \text{and} \quad \deg(\Gr(2, 6)) = \frac 1 {2880} \cdot 8! = 14
\end{equation*}
\lstinputlisting[language = python]{./gr26.sage}
\printbibliography
\end{document}

>>>>

    Here $\mathrm{lt}(I)$ is the leading term ideal of $I$, i.e. the ideal generated by the leading terms of all $f \in I$, w.r.t. $\preceq$.
\end{proposition}
\begin{proof}
    See \cite{kreuzer}.
\end{proof}
To apply this, first of all we have to collect information about the leading term ideal of $\I(\Gr(2, n))$.
This is done in the following lemma.
\begin{lemma}
    \label{prop:leading_term_ideal_grassmanian}
    Define the graded reverse monomial ordering $\preceq$ on $R := k[x_{i, j} \ | \ i < j]$ where the variables $x_{i, j}$ are ordered lexicographically w.r.t. $(i, j)$, i.e.
    \begin{equation*}
        x_{i, j} \leq x_{u, v} \ :\Leftrightarrow \ (i, j) \leq_{\text{lex}} (u, v)
    \end{equation*}
    Moreover, let
    \begin{equation*}
        I := \langle x_{i, j}x_{u, v} + x_{i, v}x_{j, u} - x_{i, u}x_{j, v} \ | \ i < j < u < v \rangle \leq R
    \end{equation*}
    be the ideal defining $\phi(\Gr(d, V))$ that was considered above. Then
    \begin{equation*}
        \mathrm{lt}(I) = J := \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle \leq R
    \end{equation*}
\end{lemma}
\begin{proof}
    We show that the set $G = \{ x_{iv} x_{ju} - x_{iu} x_{jv} + x_{ij} x_{uv} \ | \ i < j < u < v \}$ is a Gröbner basis of $I$.
    The claim then follows by the properties of Gröbner basis.

    To do this, we consider for $a < b < c < d$ and $i < j < u < v$ the polynomial
    \begin{align*}
        S := &\frac {\mathrm{lcm}(x_{iv} x_{ju}, \ x_{ad} x_{bc})} {x_{iv} x_{ju}} (x_{iv} x_{ju} - x_{iu} x_{jv} + x_{ij} x_{uv}) \\
        &- \frac {\mathrm{lcm}(x_{iv} x_{ju}, \ x_{ad} x_{bc})} {x_{ad} x_{bc}} (x_{ad} x_{bc} - x_{ac} x_{bd} + x_{ab} x_{cd})
    \end{align*}
    and show that they are reduced to $0$ by multivariate polynomial division by $G$.

    \paragraph{Case 1} If $(i, v) = (a, d)$ and $(j, u) \neq (b, c)$. Then
    \begin{equation*}
        \mathrm{lcm}(x_{iv} x_{ju}, \ x_{ad} x_{bc}) = x_{ju} x_{ad} x_{bc}
    \end{equation*}
    and so
    \begin{equation*}
        S = -x_{au} x_{jd} x_{bc} + x_{aj} x_{ud} x_{bc} + x_{ac} x_{ju} x_{bd} - x_{ab} x_{ju} x_{cd}
    \end{equation*}
    wlog assume that $j < b$, so $\mathrm{lt}(F) = -x_{au} x_{jd} x_{bc}$.
    Then there are three sub-cases.
    \subparagraph{Case 1.1} If $j < u < b < c < d$. 
    Then the only way to do cancel the leading term by polynomial division is to divide by $x_{jd} x_{bc} - x_{jc} x_{bd} + x_{jb} x_{cd}$ to get
    \begin{equation*}
        S' = x_{aj} x_{ud} x_{bc} + x_{ac} x_{ju} x_{bd} - x_{au} x_{jc} x_{bd} - x_{ab} x_{ju} x_{cd} + x_{au} x_{jb} x_{cd}
    \end{equation*}
    So $\mathrm{lt}(S') = x_{aj} x_{ud} x_{bc}$ and the only way to cancel the leading term is to divide by $x_{ud} x_{bc} - x_{uc} x_{bd} + x_{ub} x_{cd}$ to get
    \begin{equation*}
        S'' = x_{ac} x_{ju} x_{bd} - x_{au} x_{jc} x_{bd} + x_{aj} x_{uc} x_{bd} - x_{ab} x_{ju} x_{cd} + x_{au} x_{jb} x_{cd} - x_{aj} x_{ub} x_{cd}
    \end{equation*}
    Now $\mathrm{lt}(S'') = x_{ac} x_{ju} x_{bd}$ and the only way to cancel the leading term is to divide by $x_{ac} x_{ju} - x_{au} x_{jc} + x_{aj} x_{uc}$ to get
    \begin{equation*}
        S''' = -x_{ab} x_{ju} x_{cd} + x_{au} x_{jb} x_{cd} - x_{aj} x_{ub} x_{cd}
    \end{equation*}
    Now $\mathrm{lt}(S''') = -x_{ab} x_{ju} x_{cd}$ and the only way to cancel the leading term is to divide by $x_{ab} x_{ju} - x_{au} x_{jb} + x_{aj} x_{ub}$ to get $0$.
    So this subcase is finished.

    You can now do all other cases in exactly the same way, however this is quite a huge lot of effort.
    Therefore, I will not do it here.

    However, observe that $S$ only depends on the relative order of $i, j, u, v, a, b, c, d$.
    All possible orders of those are already present if we consider
    \begin{equation*}
        1 \leq i < j < u < v \leq 8 \quad \text{and} \quad 1 \leq a < b < c < d \leq 8
    \end{equation*}
    In other words, it suffices to show that $G$ is a Gröbner basis in the case $n = 8$.
    By hand, this requires exactly the same computation as started above, but it can be very easily checked using Computer Algebra.
    I have done this, and the result is as expected.
\end{proof}
At first glance, this lemma above seems quite deep, but it really is a repetitive application of the theory of Gröbner basis.

Now we have to study the structure of monomials generating $J = \langle x_{i, v} x_{j, u} \ | \ i < j < u < v \rangle$.
At first glance, this seemed to me to be much simpler than the above, but in fact it turns out not to be, as there is some subtle combinatorial structure involved.
Hence, we will first introduce the combinatorial framework in which we will work now.
\begin{definition}
    Consider the graph $G_n = (V_n, E_n)$ where $V_n = \{ (i, j) \ | \ 0 \leq i < j \leq n \}$ and
    \begin{equation*}
        E_n = \{ \{ (i, v), (j, u) \} \ | \ 0 \leq i < j < u < v \leq n \}
    \end{equation*}
    Let further
    \begin{equation*}
        s_n(d) := |\{ I \subseteq V_n \ | \ I \ \text{independent set of size $d$}\}|
    \end{equation*}
\end{definition}
It might be possible to see from the definition how independent sets in this graph are connected to monomials in $J$.
If not, the connection is explained in the proof of the next theorem.
\begin{proposition}
    Let $d$ be the largest integer such that $s_n(d) \neq 0$, i.e. the size of the largest independent set in $G_n$.
    Then
    \begin{equation*}
        \dim(\Gr(2, n)) = d - 1
    \end{equation*}
    and
    \begin{equation*}
        \deg(\Gr(2, n)) = s_n(d)
    \end{equation*}
\end{proposition}
\begin{proof}
    The idea is to apply Macaulay's basis theorem to get information about the Hilbert function of $\Gr(2, n)$.
    To this end, observe that all the monomials generating $\mathrm{lt}(\I(\Gr(2, n)))$ are square-free.
    Hence, we can partition all monomials not in $\mathrm{lt}(\I(\Gr(2, n)))$ into sets depending on which variables occur in them.
    Doing this shows that leading coefficient of the Hilbert polynomial depends only on the groups with the maximal count of monomials.
    These monomials then form a maximal independent set in $G_n$, which allows us to relate degree and dimension of $\Gr(2, n)$ to $s_n(d)$.

@@

    On the other hand, it is easy to observe that the 0-th layer
    \begin{equation*}
        I = \{ v \in V \ | \ l(v) = 0 \}
    \end{equation*}
    is indeed an independent set, as for all elements $(i, j) \in I$ have either $i = 1$ or $j = n$.
\end{proof}
\begin{corollary}
    Have $\dim(\Gr(2, n)) = 2n - 4$.
\end{corollary}
\begin{proof}
    Follows directly from the previous two statements.
\end{proof}
I also think that it is not too difficult to show that
\begin{equation*}
    s_n(2n - 3) = \frac {(2(n - 2))!} {(n - 2)! (n - 1)!}
\end{equation*}
which then shows the degree formula for the case $d = 2$.
However, this is not a lecture on graph theory or combinatorics, and I already spent much more time on this Miniproject than on the others.
So I will not do the proof here, but I think it is already very interesting to see those graphs and the connection of $\deg(\Gr(2, n))$ to combinatorial problems.
Finally, in the case of $n = 6$ we can simply count the independent sets by hand, and find the following result.
\begin{corollary}[Question (e)]
    Have $\dim(\Gr(2, 6)) = 8$ and $\deg(\Gr(2, 6)) = 14$.
\end{corollary}
\begin{proof}
    The dimension follows directly from the previous general case.
    For the degree, just count the maximal independent sets in $G_6$, as displayed in Example~\ref{ex:Gn}.
    This is slightly tricky, but it is not too hard to see that there are exactly 14 of them.
\end{proof}
Of course, the above statement on $\Gr(2, 6)$ can also easily be found using Computer Algebra.
For example, the following Sage script shows that
\begin{equation*}
    p_{\Gr(2, 6)} = \frac 1 {2880} t^8 + \frac 1 {120} t^7 + \frac {41} {480} t^6 + \frac {39} {80} t^5 + \frac {541} {320} t^4 + \frac {291} {80} t^3 + \frac {3401} {720} t^2 + \frac {101} {30} t + 1
\end{equation*}
from which we easily deduce that
\begin{equation*}
    \dim(\Gr(2, 6)) = 8 \quad \text{and} \quad \deg(\Gr(2, 6)) = \frac 1 {2880} \cdot 8! = 14
\end{equation*}
\lstinputlisting[language = python]{./gr26.sage}
\printbibliography
\end{document}
